SubGenome Analysis

### 2017-07-28 R/SubGenome Analysis.Rmd
Processing moved to parseForSubgenomes.R
Also converted to dply syntax for readability and simplified the variables needed as well as create useful outputs to compare homeologs by gene, block, and chromosome.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(readxl)
library(plyr)#
library(dplyr)
library(tidyr)
#library(reshape2)
library(ggplot2)
```
```{r preamble, echo=FALSE}
paste("This analysis is using the file", params$inputDataFile)
```

@@

```{r importSynMapData, include=FALSE}
syntelogs.raw <- read_delim(params$inputDataFile, "\t", escape_double = FALSE, trim_ws = TRUE)
#syntelogs.complete <- syntelogs.raw[complete.cases(syntelogs.raw),]

syntelogs.raw %>% 
  group_by(block, org_chr1, org_chr2) %>% 
  summarise(median_ks=median(ks, na.rm=TRUE), geneCount=n()) %>% 
  left_join(syntelogs.raw, ., by = c("block", "org_chr1", "org_chr2")) %>% 
  mutate(gene_length1=abs(stop1-start1)) %>% 
  mutate(gene_length2=abs(stop2-start2)) %>% 
  mutate(chr1=regmatches(org_chr1, regexpr("\\d*$",org_chr1))) %>% 
  mutate(chr2=regmatches(org_chr2, regexpr("\\d*$",org_chr2))) %>% 
  mutate(id=paste(block, org_chr1, org_chr2, sep="||")) -> syntelogs.mutated
```
```{r reportSynMapData, echo=FALSE, results="asis"}
# Report results of import
cat(sprintf("We have imported %d lines from the parsed SynMap file and removed %d incomplete rows resulting in %d syntenic genes.", nrow(syntelogs.raw), nrow(syntelogs.raw)- nrow(syntelogs.complete), nrow(syntelogs.complete)))
```

@@

```{r pickKSCutoff, echo=FALSE}
ks_cutoff <- 10^params$log10_ks_cutoff
qplot(log10(syntelogs.complete$ks), geom="histogram", binwidth=.01)  + geom_vline(mapping = NULL, data = NULL, xintercept = params$log10_ks_cutoff, na.rm = FALSE, show.legend = NA)
```

@@

> But how to group subgenomes over multiple chromosomes?

```{r findSubgenomes, include=FALSE}
## Find which sets of chromosomes with syntelogs should be in group 1 or group 2, where group 1 has larger syntenic blocks
# First group rows with same syntenic block and chromosome, then summarize each block's ks values with median/mean/count
syntelogs.aggregate <- ddply(syntelogs.complete, ~block+org_chr1+org_chr2, summarise, median=median(ks), mean=mean(ks), count=length(ks))

# Following the schnable article, syntenic blocks must have 12 genes "The median synonymous substitution rate of all gene pairs in a syntenic block between maize and sorghum can be used to classify syntenic blocks of 12 or more genes unambiguously as orthologous or homoeologous, however" and have a median ks value that discriminates for the alpha duplication event.
homeologs <- subset(syntelogs.aggregate, count>=12 & median<=ks_cutoff)

# Determine which chromosome has largest syntenic block, the syntelogs on this chromosome are subgenome 1
homeologs.counts <- ddply(homeologs, ~org_chr1+org_chr2, summarise, total_genes=sum(count))
homeologs.sorted <- homeologs.counts[order(homeologs.counts["org_chr1"], -homeologs.counts["total_genes"]),]

# homeologs.sorted[match(unique(homeologs.sorted$org_chr1), homeologs.sorted$org_chr1),] #first instance of each group (i.e. biggest=sub1)
homeologs.sub1rows <- match(unique(homeologs.sorted$org_chr1), homeologs.sorted$org_chr1) #not exactly right!!!! how to split sub 1 over multiple chromosomes?
homeologs.sub1 <- homeologs.sorted[homeologs.sub1rows,]
homeologs.sub2 <- homeologs.sorted[-homeologs.sub1rows,]
homeologs.sub1$subGenome="sub1"
homeologs.sub2$subGenome="sub2"

# Reconstructed ancestral chromosomes
homeologs %>% group_by(org_chr1, org_chr2) %>% summarize(genes=sum(count)) %>% spread(org_chr2, genes) -> homeologs.spread
homeologs.spread$total <- rowSums(homeologs.spread[-1], na.rm = TRUE)
syntelogs.complete %>% select(org_chr1) %>% distinct() %>% left_join(homeologs.spread) %>% replace(is.na(.), 0) %>% arrange(desc(org_chr1)) -> homeologs.totals
```
```{r reportFindSubgenomes, echo=FALSE, results="asis"}
homeologs.totals %>% knitr::kable(caption = "Table")
homeologs.totals %>% filter(total==0) %>% nrow() %>% replace(is.na(.), 0) %>% sprintf("There are %i unaligned sorghum chromosome(s) in the above table.", .) %>% cat()
```

@@

```{r writeSubgenomesToFile, echo=FALSE}
## Now recover which genes are in these sets
subgenome.groups <- rbind(merge(homeologs, homeologs.sub1, all.x = F), merge(homeologs, homeologs.sub2, all.x = F))
subgenome <- merge(syntelogs.complete, subgenome.groups, all.x = F)
write.table(subgenome, params$outputFile, sep="\t")
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```
```{r preamble, echo=FALSE}
paste("This analysis is using the file", params$inputDataFile)
```

@@

```{r importSynMapData, include=FALSE}
source("parseForSubgenomes.R")
```
```{r reportSynMapData, echo=FALSE, results="asis"}
# Report results of import
cat(sprintf("We have imported %d lines from the parsed SynMap file.", nrow(syntelogs.raw)))
```

@@

```{r pickKSCutoff, echo=FALSE}
qplot(log10(syntelogs.complete$ks), geom="histogram", binwidth=.01)  + geom_vline(mapping = NULL, data = NULL, xintercept = params$log10_ks_cutoff, na.rm = FALSE, show.legend = NA)
```

@@

```{r reportFindSubgenomes, echo=FALSE, results="asis"}
## Display chromosomes with significant syntenic genes between org1 and org2
homeologs.sorted %>%
  select(org_chr1, org_chr2, chromosomeGeneCount) %>%
  spread(org_chr2, chromosomeGeneCount) -> homeologs.spread

homeologs.spread$total <- rowSums(homeologs.spread[-1], na.rm = TRUE)

homeologs.spread %>% knitr::kable(caption = "Table")
homeologs.spread %>% filter(total==0) %>% nrow() %>% replace(is.na(.), 0) %>% sprintf("There are %i unaligned sorghum chromosome(s) in the above table.", .) %>% cat()
```

@@

```{r writeSubgenomesToFile, echo=FALSE}
## Now recover which genes are in these sets
write.table(subgenome, params$outputFile, sep="\t")
```
